Individual heterogeneity in the functional network topography of medial prefrontal cortex

### 2019-05-21 Manuscript.Rmd
Done with the methods

All network analyses were performed using the igraph package [v. 1.1.2; https://igraph.org/r/; @G.2006] in R [v. 3.4.1; https://www.r-project.org/; @RTeam2018]. To establish each subject's network, we selected all the vertices contained within the mPFC/PCC search space (n = 4,801 per subject; mPFC = 2854, PCC = 1947) and computed the Pearson correlation of the time series for every pair of vertices. This produced a weighted network for each subject, in which the nodes were surface vertices and the edges were the correlations among them. Next, we applied Fisher's r to z transformation, and performed a two-sided significance test to identify significant connections. The resulting p-values were corrected for multiple comparisons (FDR < 0.05), and edges with non-significant correlations were set to 0 (all other edges retained their respective z-values). The proportion of remaining edges (i.e. network density) was high for all individuals (mean = `r round(mdAll, digits = 2)`, SD = `r round(sddAll, digits = 3)`), and the retained edges mostly consisted of positive correlations (mean proportion positive = `r mPos`, sd = `r sdPos`). Next, we took the exponential of these values so that all weights were positive while maintaining the ordinal ranks of the original correlation distribution within the set of retained edges. Non-retained edges were reset to 0 to exclude them from the network analyses. We generated and analyzed network weight matrices at four levels: (1) for each subject's full dataset (4800 TRs total); (2) on each step of a sliding window analysis (see Partition Evaluation for more details); (3) for the concatenated time series for the two runs on each day (2400 TRs total); and (4) for each run separately (1200 TRs each).

@@

Where $I$ is an identity matrix of size $n$, and $D$ is a diagonal matrix containing the strength of each vertex (i.e. the sum of its edge weights with all other vertices). This resulted in a matrix wherein each entry was the negative normalized value of the connection (from 0 to 1) between any two vertices relative to their combined connectivity strength, and with ones along the diagonal. The transformation ensures that every row sums to zero. We then computed the eigenvalues and eigenvectors of the symmetric normalized Laplacian matrix, and used the eigenvector associated with the second-to-lowest eigenvalue to divide the network into two. This vector (traditionally called the 'Fiedler vector') provides a set of positive and negative values and is binarized by sign [@Fiedler1975]. This method partitions the network into two similarly-sized communities. In this way, SP avoids producing communities that are too small to be physiologically meaningful (for example, small sets of vertices that are spuriously correlated due to measurement noise). Given that this data-driven method does not label the two communities or establish correspondence across participants, we defined each individual's "DN" community as that which contained the DN-specific PCC parcel identified in our meta-analysis. The high density of the graphs ensured that spectral partitioning did not face the issues associated with its use in sparse networks [@Fortunato2016].

In order to evaluate the validity of the resulting partitionings across community-detection methods, we also estimated network communities using the more traditional approach of modularity maximization [@Garcia2017], based on the algorithm from Clauset et al. [-@Clauset2004]. This method compares the strength of the connection between any two vertices to their probability of being randomly connected to any other vertex in the network. The method heuristically iterates through many possible combinations of vertices, and selects the partitioning that maximizes the within-community edge weights, relative to a random network containing the same number of edges. Unlike SP, modularity can fractionate a network into more than two communities. Agreement between the partitions provided by the bounded and unbounded community detection methods would suggest the results are not distorted by the restriction of SP to binary partitionings.

@@

We used the adjusted Rand index (ARI) to evaluate the stability and topographical heterogeneity of the communities within and across individuals [@Hubert1985], which was calculated using the "mcclust" package in R [@Arno2012]. The base of the ARI is computed by the formula

@@

where $a$ is the number of pairs of nodes that were grouped together in both partitionings, $b$ is the number that were grouped separately, and $c$ and $d$ denote the number of nodes grouped together (separately) in one partitioning, but separately (together) in the other. Therefore, the ARI estimates the fraction of all possible node pairs that had the same status (connected or not) in both partitionings (with the denominator equal to $n(n - 1)/2$). The resulting ratio is adjusted against a baseline given by the expectation assuming independent partitionings. The index ranges from 0 to 1, where 0 denotes the value expected by chance. This means that even though clustering deviations are heavily penalized, positive ARI values compare favorably against chance clustering (and can take negative values if the index given by the formula above falls below the chance level). In short, the ARI determines the chance-corrected agreement between any two partitions while being agnostic to the labeling scheme.

We performed a number of comparisons among partitions. First, we computed the degree of agreement between SP and modularity per subject. SP and Modularity have been previously found to show underfitting and overfitting tendencies, respectively, in their community detection performance in a diverse number of network types [@Ghasemian2018], so alignment between the two algorithms would increase our confidence in the validity of the resulting partitionings. Next, we compared the subject-level SP partitionings across individuals, and calculated the mean pairwise ARI for the group. We then performed the same evaluation for PCC and mPFC separately, and examined whether there were mean differences in overall agreement within these regions by performing a paired permutation analysis (5000 iterations) on all pairwise ARI comparisons across subjects. Given that the functional heterogeneity of brain areas is unevenly distributed across the cortex [@Mueller2013].

To estimate the degree of stability of partitionings per individual we performed a sliding window analysis (20 min windows, 1 min increments, median number of windows per subject = 37, range = 35 - 37), comparing each window's resulting partitioning against the partitioning derived from the whole subject's data. A 20-min window is typically adequate for identifying stable community features in brain networks [@Gordon2017]. We assessed whether the magnitude of the Fiedler vector value for a given vertex (for the full subject-level data set) was associated with the stability of that vertex’s sub-network assignment across time windows. To do this, we fit a mixed effects logistic regression model, in which the dependent variable was the proportion of times each vertex participated in the DN community across windows, and the explanatory variables included a random effect of subject and a fixed effect of the Fiedler vector value for that vertex (derived from their full time series). Based on this relationship, we identified a threshold Fiedler vector value for each subject, such that above-threshold vertices were persistently associated with either DN or non-DN more than 99% of the time.

We then estimated the level of agreement between network partitions estimated using data across individual scan days (with 2 days per participant). If the functional organization estimated by SP is indeed individual-specific, we should see higher agreement within individual (test/re-test across days) than across individuals. We tested this idea by computing the ratio of the mean ARI within and between individuals. Ratios close to one denoted similar within-participant and across-participant alignment, whereas ratios considerably higher than one suggested that partitions were more similar within-participant than across participants. We then extended this idea by computing the agreement across individual runs (4 per subject). Similar to the day-based analysis, we assessed whether run-level data showed higher agreement within-subject than between subjects.

@@

We evaluated the performance of the SP algorithm in comparison to a simpler partitioning approach using seed-based functional connectivity. First, we estimated each subject's DN partition in mPFC based on its vertex-wise functional correlation (Pearson) with the average activity across all vertices in PCC for each day separately. We compared these seed-based maps with the unthresholded Fiedler vectors produced by SP across days in three ways: 1) between seed-based maps; 2) between SP-based maps; and 3) between seed- and SP-based maps. Because the values in the maps were continuous-valued (and not categorical labels, which would be amenable to ARI), we quantified the similarity between maps in terms of the spatial Spearman correlation across vertices. These spatial correlations were meant to determine the test/re-test reliability of each approach, as well as the overall level of agreement between them. For 3 subjects, the maps produced with one of the days' datasets insufficiently covered area 7m, and so the community labeling was reversed (for one additional subject, both daily maps were reverted). ARI is robust to these labeling issues, but this produced strong negative correlations of the Fiedler vector for these subjects, even though the topography was highly similar. We thus visually inspected these subjects (n = 7) and reverted the labeling when it was clear that the topologies aligned across days.

>>>>

All network analyses were performed using the igraph package [v. 1.1.2; https://igraph.org/r/; @G.2006] in R [v. 3.4.1; https://www.r-project.org/; @RTeam2018]. To establish each subject's network, we selected all the vertices contained within the mPFC/PCC search space (n = 4,801 per subject; mPFC = 2854, PCC = 1947) and computed the Pearson correlation of the time series for every pair of vertices. This produced a weighted network for each subject, in which the nodes were surface vertices and the edges were the correlations among them. Next, we applied Fisher's r to z transformation, and performed a two-sided significance test to identify significant connections. The resulting p-values were corrected for multiple comparisons (FDR < 0.05), and edges with non-significant correlations were set to 0 (all other edges retained their respective z-values). The proportion of remaining edges (i.e. network density) was high for all individuals (mean = `r round(mdAll, digits = 2)`, SD = `r round(sddAll, digits = 3)`), and the retained edges mostly consisted of positive correlations (mean proportion positive = `r mPos`, sd = `r sdPos`). Next, we took the exponential of these values so that all weights were positive while maintaining the ordinal ranks of the original correlation distribution within the set of retained edges. Non-retained edges were reset to 0 to exclude them from the network analyses. We generated and analyzed network weight matrices at four levels: (1) for each subject's full concatenated dataset (up to 4800 TRs); (2) on each step of a sliding window analysis (see Partition Evaluation for more details); (3) for the concatenated time series for the two runs on each day (up to 2400 TRs); and (4) for each run separately (up to 1200 TRs).

@@

Where $I$ is an identity matrix of size $n$, and $D$ is a diagonal matrix containing the strength of each vertex (i.e. the sum of its edge weights with all other vertices). This resulted in a matrix wherein each entry was the negative normalized value of the connection (from 0 to 1) between any two vertices relative to their combined connectivity strength, and with ones along the diagonal. The transformation ensures that every row sums to zero. We then computed the eigenvalues and eigenvectors of the symmetric normalized Laplacian matrix, and used the eigenvector associated with the second-to-lowest eigenvalue (traditionally called the 'Fiedler vector') to divide the network into two. The Fiedler vector provides a set of positive and negative values and is binarized by sign to partition the network into two similarly-sized communities [@Fiedler1975]. In this way, SP avoids producing communities that are too small to be physiologically meaningful (for example, small sets of vertices that are spuriously correlated due to measurement noise). Given that this data-driven method does not label the two communities or establish correspondence across participants, we defined each individual's "DN" community as that which contained the DN-specific PCC parcel identified in our meta-analysis. The high density of the graphs ensured that SP did not face the issues associated with its use in sparse networks [@Fortunato2016].

In order to evaluate the validity of the resulting partitionings across community-detection methods, we also estimated network communities using the more traditional approach of modularity maximization [@Garcia2017], based on the algorithm from Clauset et al. [-@Clauset2004]. The method heuristically iterates through many possible combinations of vertices, and selects the partitioning that maximizes the within-community edge weights, relative to a random network containing the same number of edges and communities. Unlike SP, modularity can fractionate a network into more than two communities. Agreement between the partitions provided by the bounded and unbounded community detection methods would suggest the results are not distorted by the restriction of SP to binary partitionings.

@@

We used the adjusted Rand index (ARI) to evaluate the stability and topographical heterogeneity of the communities within and across individuals [@Hubert1985], which was calculated using the "mcclust" package in R [@Arno2012]. The ARI is a metric that quantifies the similarity between two alternative clusterings of the same data. The base of the ARI is computed by the formula

@@

where $a$ is the number of pairs of nodes that were grouped together in both partitionings, $b$ is the number that were grouped separately, and $c$ and $d$ denote the number of nodes grouped together (separately) in one partitioning, but separately (together) in the other. Therefore, the ARI estimates the fraction of all possible node pairs that had the same status (connected or not) in both partitionings (with the denominator equal to $n(n - 1)/2$). The resulting ratio is adjusted against a baseline given by the expectation assuming independent partitionings to yield an index that ranges from 0 to 1, where 0 denotes the value expected by chance. This means that even though differences are heavily penalized, positive ARI values compare favorably against chance clustering (and can take negative values if the index given by the formula above falls below the chance level). In short, the ARI quantifies the chance-corrected agreement between any two partitions while being agnostic to the labeling scheme.

We performed a number of comparisons among partitions. First, we computed the degree of agreement between SP and modularity per subject. SP and Modularity have been previously found to show underfitting and overfitting tendencies, respectively, in their community detection performance in a diverse set of network types [@Ghasemian2018], so alignment between the two algorithms would increase our confidence in the validity of the resulting partitionings. Next, we compared the subject-level SP partitionings across individuals, and calculated the mean pairwise ARI for the group. We then performed the same evaluation for PCC and mPFC separately, and examined whether there were mean differences in overall agreement within these regions by performing a paired permutation analysis (5000 iterations) on all pairwise ARI comparisons across subjects.

To estimate the degree of variability of partitionings per individual we performed a sliding window analysis (20 min windows, 1 min increments, median number of windows per subject = 37, range = 35 - 37), comparing each window's resulting partitioning against the partitioning derived from the subject's whole data set. A 20-min window has previously been found adequate for identifying stable community features in brain networks [@Gordon2017]. We assessed whether the magnitude of the Fiedler vector value for a given vertex (for the full subject-level data set) was associated with the stability of that vertex’s sub-network assignment across time windows. To do this, we fit a mixed effects logistic regression model, in which the dependent variable was the proportion of times each vertex participated in the DN community across windows, and the explanatory variables included a random effect of subject and a fixed effect of the Fiedler vector value for that vertex (derived from their full time series). Based on this relationship, we identified a threshold Fiedler vector value for each subject, such that above-threshold vertices were persistently associated with either DN or non-DN more than 99% of the time.

We then estimated the level of agreement between network partitions estimated using data across individual scan days (with 2 days per participant). If the functional organization estimated by SP is indeed individual-specific, we should see higher agreement within individual (test/re-test across days) than across individuals. We tested this idea by computing the ratio of the mean ARI within and between individuals. Ratios close to one would denote similar within-participant and across-participant alignment, whereas ratios considerably higher than one would suggest that partitions were more similar within-participant than across participants. We then extended this idea by computing the agreement across individual runs (4 per subject). Similar to the day-based analysis, we assessed whether run-level data showed higher agreement within-subject than between subjects.

@@

We evaluated the performance of the SP algorithm in comparison to a simpler partitioning approach based on seed-based functional connectivity. For each day (2 per individual), we we estimated each subject's DN partition in mPFC based on its vertex-wise functional correlations (Pearson) with the spatially averaged activity across all vertices in PCC. The sign of the Fiedler vector was reoriented so the DN community was marked by positive values in every subject. We compared these seed-based maps with the unthresholded Fiedler vectors produced by SP by calculating three sets of across-day similarity values for each individual: 1) between the two seed-based maps; 2) between the two SP-based maps; and 3) between seed- and SP-based maps. Because the values in the maps were continuous-valued (and not categorical labels, which would be amenable to ARI), we quantified the similarity between maps in terms of the spatial Spearman correlation across vertices. These spatial correlations were meant to determine the test/re-test reliability of each approach, as well as the overall level of agreement between them. For 7 subjects, the community maps produced with one of the days' data sets insufficiently covered area 7m (i.e. less than 50% coverage), and so the sign of the Fiedler vector did not match the community assignment specified above. ARI is robust to such labeling issues, but this produced strong negative correlations of the Fiedler vector for these subjects across days, even though the topography was highly similar. We thus visually inspected these subjects and reverted the labeling when it was clear that the topography aligned across days.
